Zohmg---a large scale data store for aggregated time-series-based data. [DRAFT]

### 2009-05-08 doc/report/msc-report.tex
rounded out introduction, merged subsection on data size.

In this thesis we present a data store called Zohmg. Zohmg is built on top of
HBase and uses Hadoop to run import jobs in parallel. It solves the problem of
summing, storing and serving measurements. These measurements may be taken
from log files or similar sources. Zohmg stores only the summarized data, not
the atomic measurements. The data store is optimized for speed of data
retrieval: one of the design goals was to serve data at mouse-click rate and
promote real-time data exploration.

@@

Inevitably, these services and operations will generate a log for later
analysis. As the services grow in size, so will the data sets they generate grow.

An important property of these data sets is that they generally have a regular
structure and are represented in independent records. The records can be
broken up without losing any of their meaning and are as such very well suited
for parallel analysis. The data sets are not fit to store using traditional
database techniques, if only for the fact that they are too large to fit in a
relational database.

@@

\subsection{Data size}

The sheer size of the data makes it unfeasible to use a single computer.
Eventually a single machine would be fed so much data that it cannot cope
with it. For instance a single web log file for one day can be up to
several gigabytes. It is problematic for a single machine to handle this
because of the limitations of memory, loading the entire file in memory.
Distributed computing solves this by spreading the computations out to
several machines in parallel. The obvious gain is that the waiting time
before a computation is done is divided by the amount of nodes used,
compared to running the computation on one machine. There is also a gain
related to the actual machines, it is possible to use commodity hardware.
Because MapReduce and BigTable has high availability and reliability a node
in the cluster can safely be replaced on the fly if it fails.

Scalability is also very important. Using single machines for computing is
problematic when the data set grows, more powerful machines has to be
obtained. More power contained in a single machine results in higher price,
eventually it is too expensive to perform computations on very large
data sets. Both MapReduce and BigTable scale close to linearly, adding one
machine to the cluster adds equal amount of processing power. Since
MapReduce and BigTable can use commodity hardware it is reasonable to buy
extra machines to gain extra computing power or storage space.

>>>>

We present a data store called Zohmg. It is built on top of HBase and uses
Hadoop to run import jobs in parallel. Zohmg stores and serves aggregates of
measurements. The measurements are extracted taken from log files or similar
sources with user-supplied code.

The data store is optimized for speed of data retrieval: one of the design
goals was to serve data at mouse-click rate and promote real-time data
exploration. As such, Zohmg stores only the summarized data, not the atomic
measurements.

@@

Inevitably, these services and operations will generate a log for later
analysis. As the services grow in size, so will the data sets they generate.

An important property of these data sets is that they generally have a regular
structure and are represented as a number of independent records. The records
can be broken up without losing any of their meaning and are as such very well
suited for parallel analysis. The data sets are not fit to store using
traditional database techniques, if only for the fact that they are too large
to fit in a relational database.

Each log record typically contains quite a bit of information. For example, a
log of visits to a web site might contain a timestamp, information about what
country the visiting user is from, and a string of text that identifies the
web browser that was used. From these base facts one can collect information
about what countries generate the most traffic to the site, and which web
browser is most popular in a certain country, for example.

These bits of information can be thought of as points along an axis; Sweden,
Norway and Denmark are all points along the country axis, for instance. This
is an important concept and it will be used throughout the text.

Typically, the size of a data set is in the range of several gigabytes of data
per day. A single computer is not likely to be able to process that amount of
data before a new batch arrives, and for very large data sets the single
computer would simply not be able to fit the intermediate data set in its
memory at all. Some data sets are many times larger than that and do not even
fit inside a single computers storage devices.

It is suitable to store the data across many storage nodes in what is called a
distributed filesystem and to to exploit the parallel nature of the records by
breaking up the data sets in pieces and have each piece be processed
separately on a node in a large compute cluster.

% TODO: fixa cites. Luckily, this is a problem that has already been solved.
Google has published papers on the MapReduce framework \cite{mapreduce}, which
allows for parallel processing of large data sets, and The Google Filesystem
\cite{gfs}, a scalable, distributed filesystem. The Apache Hadoop project has
published open source implementations of a mapreduce framework and a
distributed filesystem, both of which build on the ideas in the papers
published by Google.

@@
